A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-09-25 appendix.Rmd
remove unused/commented parts

```{r}
# OOP supports needed by `visage`
# remotes::install_github("TengMCing/bandicoot")
# 
# Visual inference models and p-value calculation
# remotes::install_github("TengMCing/visage")

>>>>

```{r}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# OOP supports needed by `visage`
if (!requireNamespace("bandicoot", quietly = TRUE)) {
  remotes::install_github("TengMCing/bandicoot")
}

# Visual inference models and p-value calculation
if (!requireNamespace("visage", quietly = TRUE)) {
  remotes::install_url("https://github.com/TengMCing/visage/raw/master/built/visage_0.1.2.tar.gz")
} else {
  if (packageVersion("visage") < "0.1.2") {
    remotes::install_url("https://github.com/TengMCing/visage/raw/master/built/visage_0.1.2.tar.gz")
  }
}

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

if (!requireNamespace("here", quietly = TRUE)) {
  install.packages("here")
}

if (!requireNamespace("kableExtra", quietly = TRUE)) {
  install.packages("kableExtra")
}

if (!requireNamespace("ggpubr", quietly = TRUE)) {
  install.packages("ggpubr")
}

if (!requireNamespace("lvplot", quietly = TRUE)) {
  install.packages("lvplot")
}

### 2023-09-25 paper.Rmd
remove unused/commented parts

```{r}
# OOP supports needed by `visage`
# remotes::install_github("TengMCing/bandicoot")
# 
# Visual inference models and p-value calculation
# remotes::install_url("https://github.com/TengMCing/visage/raw/master/built/visage_0.1.0.tar.gz")

@@

<!--- ET: I suggest removing this to save space and we don't use this plot type.

\textcolor{red}{There are also other types of plots that can be used for assessing residuals. The normal probability plot is used to assess a normality assumption on the error distribution. Residuals are often plotted against  order of observations to check for collection bias, batch effects or autocorrelation. Partial residual plots assess the importance of an independent variable to a model, conditional on other variables, by plotting partial residuals against this variable. Partial residuals are computed from the difference between the response value and the fitted model excluding term of the independent variable of interest. A partial regression plot assesses the unique effect of an independent variable in a multiple regression model, by plotting residuals from the model with and without the this variable against each other.}

--->

<!--other types of residual plot to extract the available information and identify inappropriate assumptions, like the added variable plot (cite here) for studying the role of an external variable if it enters linearly into a model. Partial residual plots (cite here) have been suggested as computationally convenient substitutes for the added variable plots. For use in diagnostic procedures, several transformations of the ordinary residuals have been suggested to overcome partially some of their shortcomings. Studentized residuals (cite here) have null distributions that are independent of the scale parameters. These residuals are shown to be closely related to a mean shift model for outliers and to the residuals obtained when each case in turn is left out of the data. Alternatively. the residuals can be transformed to have a selected covariance structure: The usual suggestion is to obtain residuals with uncorrelated elements.}-->

<!--ET: I think we need to remove this to reduce pages.

These are all topics explained thoroughly in the sizeable literature on diagnostic procedures for linear models, e.g. @draper1998applied, @montgomery1982introduction, @belsley1980regression, @cook1999applied and @cook1982residuals. Interestingly, despite the abundance of conventional tests, _all_ of these writings advise that plotting residuals is an essential tool for diagnosing regression model problems:

> *Some most useful checks allowed by data plots should be done on a routine basis for every regression.* [@draper1998applied]

> *Such formal and informal procedures are complementary, and both have a place in residual analysis.* [@cook1982residuals]

> *In our experience, statistical tests on regression model residuals are not widely used. In most practical situations the residual plots are more informative than the corresponding tests. However, since residual plots do require skill and experience to interpret, the statistical tests may occasionally prove useful.* [@montgomery1982introduction]

-->

@@

<!--that is addressed by the results in this paper because from a practical perspective we would only want to flag the model as problematic when ther is a XXX diagnostic  practical significance. These can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers and slight departures. -->

<!--
As discussed in Section \ref{conventionally-testing-for-departures}, many conventional tests are available for detecting residual departures. Implementation-wise, the built-in R package `stats` provides some commonly used residual-based tests, such as SW test. A more comprehensive collection of regression diagnostics tests can be found in the R package `lmtest` [@lmtest]. In terms of heteroskedasticity diagnostics, the R package `skedastic` [@skedastic] collects and implements 25 existing conventional tests published since 1961.

We pick RESET test (`resettest`) and BP test (`bptest`) from the R package `lmtest`, and SW test (`shapiro.test`) from the built-in R package `stats`. Among them, RESET test is the only exact and appropriate test in this scenario. Both the BP test and the SW test are approximate and inappropriate tests. Their estimated power is shown in Figure \ref{fig:polypower}. To set up the RESET test, we include different powers of fitted values as proxies. According to @ramsey1969tests, there are no general rules for the power of the fitted values needed by the RESET test, but it finds power up to four is usually sufficient. Thus, we follow this guideline to conduct the RESET test. For the BP test, the choice of predictors in the auxiliary regression is left to the user [@breusch1979simple]. But as @waldman1983note suggested, it is a good choice for the set of auxiliary predictors in the BP test be the same as the White test. Thus, we include both $x$ and $x^2$ in the auxiliary regression.
-->

@@

<!--- ET: interesting history, but not pertinent to the story

According to @abramowitz1964handbook, Hermite polynomials were initially defined by @de1820theorie, but named after @hermite1864nouveau because of the unrecognisable form of Laplace's work. 
--->

@@

<!--- ET: let's keep the language simpler.

Since $z = O(x^j)$, for $j > 1$, $z$ is a higher order term left out of the null regression, it leads to model misspecification. 
--->

@@

<!----

Figure \ref{fig:example-poly-lineup} is one of the non-linearity lineups used in experiment. This data plot is produced by the non-linearity model with $j = 6$, and is in location $2^3+1$. All five participants who viewed the lineup chose this plot as most different, thus all detected the data plot.

```{r example-poly-lineup, fig.pos="t!", fig.height = 7, fig.width = 7, fig.cap = "One of the lineups containing non-linearity patterns used in data collection period I. Can you spot the most different plot? The data plot is positioned at $2^3 + 1$."}
VI_MODEL$plot_lineup(vi_lineup$poly_24$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(),
                     remove_axis = TRUE)
```

--->

@@

<!---

```{r example-heter-lineup, fig.pos="t!", fig.height = 7, fig.width = 7, fig.cap = "One of the lineups containing heteroskedasticity pattern used in data collection period II. Can you spot the most different plot? The data plot is positioned at $3^3 - 3^2$."}

VI_MODEL$plot_lineup(vi_lineup$heter_471$data, 
                     remove_grid_line = TRUE, 
                     theme = theme_light(),
                     remove_axis = TRUE)
```

An example lineup of this model used in data collection period II is shown in Figure \ref{fig:example-heter-lineup} with $a = -1$. The data plot location is in position $2^4 + 2$. Nine out of 11 participants chose this plot from the lineup.

--->

@@

<!--- ET: previous reference was "normal" but now Gaussian. This doesn't add to the story so I suggest just remove.

The distribution may be relatively Gaussian, reflecting a linear combination of many predictors, adhering to the Central Limit Theorem.

--->

@@

<!--As mentioned in Section \ref{power-of-the-tests}, $\alpha$ used in Equation \ref{eq:pvalue-beta-binomial} needs to be estimated using null lineups. Three replications are made for $3 \times 4 = 12$ combinations of common factors $n$ and fitted value distribution, results in $12 \times 3 = 36$ lineups included in data collection period III. In these lineups, the data of the data plot is generated from a model with zero effect size, while the data of the 19 null plots are generated using the same simulation method discussed in Section \ref{visual-test-procedure-based-on-lineups}. This generation procedure differs from the canonical Rorschach lineup procedure, which requires that all 20 plots are generated directly from the null model. However, these lineups serve the same fundamental purpose: to assess the number of visually interesting plots generated under $H_0$.

To account for the fact that our simulation method for these lineups is not the Rorschach procedure, we use the method suggested in @vanderplas2021statistical for typical lineups containing a data plot to estimate $\alpha$. 
(We have included a sensitivity analysis in the Appendix to examine the impact of the variance of the $\alpha$ estimate on our findings.)

All lineups consist of only null plots are planned to be evaluated by 20 participants. 
However, presenting only these lineups to participants are considered to be bad practices as participants will lose interest quickly. 
Therefore, we plan to collect 6 more evaluations on the 279 lineups with uniform fitted value distribution, result in $(36 \times 20 + 279 \times 3 \times 6) / (20-2) = 133$ participants recruited for data collection period III.
-->

@@

In each lineup evaluation, the participant was asked to select one or more plots that are most different from others, provide a reason for their selections, and evaluate how different they think the selected plots are from others. If there is no noticeable difference between plots in a lineup, participants had the option to select zero plots without the need to provide a reason. During the process of recording the responses, a zero selection was considered to be equivalent to selecting all 20 plots. No participant was shown the same lineup twice. Information about preferred pronouns, age group, education, and previous experience in visual experiments were also collected. A participant's submission was only included in the analysis if the data plot is identified for at least one attention check. <!--Data of rejected submissions are discarded automatically to maintain the overall data quality.-->

@@

# 106
# 112
# 113
# 115
# 132
# 133
# 144!
# 147
set.seed(132)

@@

# 10091
# 101
# 105
# 106
# 112
# 119
# 133
set.seed(135)

@@

<!-- In terms of visual tests, we do not observe significant difference between the power curve of the uniform distribution and the power curve of the discrete uniform distribution. Normal and lognormal distributions produce lower power than uniform and discrete uniform distributions in regards of residual departure patterns, but the gap is larger when the residual plot contains heteroskedasticity pattern. This suggests that residual departures displayed with evenly distributed fitted values are visually more attractive and effective. Besides, the skewness of the fitted value distribution plays a role in the degree of presence of visual features as lognormal distribution produce the lowest power for both non-linearity and heteroskedasticity patterns. A symmetric fitted value distribution can better reveal the underlying shape.-->

@@

<!-- Although it is unusual to consider fitted value distribution or predictor distribution in power analysis, it is not so surprised that the power of conventional tests vary because of fitted value distribution. For instance, the power of the RESET test which is effectively a F test depends on the realizations of the additional predictors included in the auxiliary regression equation [@jamshidian2007study; @olvera2019relationship; @zhang2018practical]. And those additional predictors are linear transformation of random vectors following certain predictor distributions. What is truly unexpected is the huge variability in power of conventional tests compared to visual tests. The discrete uniform distribution constantly produce the highest power while the normal distribution and the lognormal distribution produce the lowest power for non-linearity and heteroskedasticity pattern respectively. The difference in power corresponding to different distributions peak at around 90% when the effect size is at a moderate level. This huge difference is undesirable as is not uncommon for analysts to omit the factor of predictor distribution in power calculation and sample size calculation. -->

@@

One of the primary limitations of the lineup protocol lies in its reliance on human judgements. In this context, the effectiveness of a single lineup evaluation can be dependent on the perceptual ability and visual skills of the individual. However, when results from multiple individuals are combined the outcome is encouragingly high-quality and robust. For simple plots and strong patterns just a few individuals are needed to arrive at a clear answer, but more individuals will be needed when the plot design is complex, or the signal strength is weak. <!--This reliance on human judgment introduces an element of variability and subjectivity that can potentially impact the robustness of the results obtained. To ensure reliable outcomes, it's often necessary to include a diverse group of participants capable of offering unbiased judgements. This requirement has led researchers, including those in our study, to turn to crowd-sourcing platforms that offer access to a random selection of participants, which can help mitigate potential bias.-->

@@

<!--the process of constructing and evaluating lineups becomes accessible and straightforward. This accessibility empowers a wider range of researchers, even those without extensive statistical backgrounds, to employ the lineup protocol in their studies. Moreover, the cost-effectiveness of lineup-based analyses makes them particularly appealing for integration into statistical education within classroom settings (cite adam loy paper here). The practicality of lineup demonstrations can facilitate a deeper understanding of the concept of randomness and statistical testing among students.-->

<!--However, it's important to understand that if the goal is to attain results similar to those of formal hypothesis testing in rigor and reliability, using crowd-sourcing becomes necessary. The expense associated with crowd-sourcing is relatively reasonable when compared to the costs of other laboratory devices, making it a sensible investment for research that requires clear and valuable findings. Importantly, this higher level of scrutiny is typically justified only when the data plot can not be confidently identified and requires a thorough investigation.-->

@@

<!-- Additionally, there is the exciting possibility of utilizing computer vision technology to automate the analysis of lineup of residual plots. This innovation has the potential to significantly reduce the expenses associated with recruiting human judges. By incorporating computer vision models, researchers can simplify the evaluation process, making it more efficient and cost-effective.

The scope of this study is to highlight practical value of residual plots in model diagnostics, drawing comparisons with conventional tests. Our primary focus is on the use of residual-fitted plots in a controlled model environment for the purpose of a human subject experiment. It's important to note that there are other intriguing aspects of residual diagnostics that fall beyond the boundaries of this paper. For example, as mentioned in Introduction, residual plots can be used to explore interesting data characteristics while conducting model diagnostics. Furthermore, some model violations, such as serial correlation, necessitate the use of alternative residual plots like residual against row number plots. Nevertheless, we believe that the insights derived from this study have the potential to apply to other residual diagnostics setting, because the fundamental principles and statistical concepts underlying the lineup protocol and residual plots remain consistent across different contexts. In real-world situations, researchers often rely on visual assessment to make informed decisions about model validity. The lineup protocol's capability to identify deviations from randomness serves as a valuable aid in this scenario.-->

@@

<!-- Make a note about "arguing" that there is no problem apparent in residual plot - reading a single plot -  vs the visual test which removes the argument, and provides objectivity.-->

@@

Human evaluation of residuals is expensive, time-consuming and laborious. This is possibly why residual plot analysis is often not done in practice. However, with the emergence of effective computer vision, it is hoped this work helps to lay the foundation for automated residual plot assessment. <!--The findings suggest the strong demand of graphical inspection in regression diagnostics, so developing an automatic visual inference system to evaluate lineups of residual plots is valuable. We plan to build a completed open-source system in an `R` package and provide a web interface such as a website for public to interact with. Details about this system will be discussed in our next paper.-->

>>>>

```{r}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# OOP supports needed by `visage`
if (!requireNamespace("bandicoot", quietly = TRUE)) {
  remotes::install_github("TengMCing/bandicoot")
}

# Visual inference models and p-value calculation
if (!requireNamespace("visage", quietly = TRUE)) {
  remotes::install_url("https://github.com/TengMCing/visage/raw/master/built/visage_0.1.2.tar.gz")
} else {
  if (packageVersion("visage") < "0.1.2") {
    remotes::install_url("https://github.com/TengMCing/visage/raw/master/built/visage_0.1.2.tar.gz")
  }
}

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

if (!requireNamespace("here", quietly = TRUE)) {
  install.packages("here")
}

if (!requireNamespace("ggmosaic", quietly = TRUE)) {
  install.packages("ggmosaic")
}

if (!requireNamespace("rcartocolor", quietly = TRUE)) {
  install.packages("rcartocolor")
}

if (!requireNamespace("kableExtra", quietly = TRUE)) {
  install.packages("kableExtra")
}

if (!requireNamespace("patchwork", quietly = TRUE)) {
  install.packages("patchwork")
}

@@

@@

@@

@@

@@

@@

@@

@@

@@

In each lineup evaluation, the participant was asked to select one or more plots that are most different from others, provide a reason for their selections, and evaluate how different they think the selected plots are from others. If there is no noticeable difference between plots in a lineup, participants had the option to select zero plots without the need to provide a reason. During the process of recording the responses, a zero selection was considered to be equivalent to selecting all 20 plots. No participant was shown the same lineup twice. Information about preferred pronouns, age group, education, and previous experience in visual experiments were also collected. A participant's submission was only included in the analysis if the data plot is identified for at least one attention check.

@@

set.seed(132)

@@

set.seed(135)

@@

@@

@@

One of the primary limitations of the lineup protocol lies in its reliance on human judgements. In this context, the effectiveness of a single lineup evaluation can be dependent on the perceptual ability and visual skills of the individual. However, when results from multiple individuals are combined the outcome is encouragingly high-quality and robust. For simple plots and strong patterns just a few individuals are needed to arrive at a clear answer, but more individuals will be needed when the plot design is complex, or the signal strength is weak.

@@

@@

@@

@@

Human evaluation of residuals is expensive, time-consuming and laborious. This is possibly why residual plot analysis is often not done in practice. However, with the emergence of effective computer vision, it is hoped this work helps to lay the foundation for automated residual plot assessment.

### 2023-09-25 paper.tex
remove unused/commented parts

@@

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot $2^2 + 2$, exhibiting non-linearity) is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot $2^2 + 2$.}\label{fig:first-example-lineup}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/nonlinearheterpower-1}

\caption{Comparison of power between different tests for (A) non-linear and (B) heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating non-reject and reject in visual testing of lineups. The multiple lines for the visual test arise from estimating the power on many bootstrap samples. The row of scatterplots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot.}\label{fig:nonlinearheterpower}
\end{figure}

@@

One of the primary limitations of the lineup protocol lies in its
reliance on human judgements. In this context, the effectiveness of a
single lineup evaluation is closely connected with the perceptual
ability and visual skills of the individual. When results from multiple
individuals are combined the outcome is encouragingly robust. For simple
plots and strong patterns just a few individuals are needed to arrive at
a clear answer, but more individuals will be needed when the plot design
is complex, or the signal strength is weak.

>>>>

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting the data plot (at position $2^2 + 2$, exhibiting non-linearity).}\label{fig:first-example-lineup}
\end{figure}

@@

@@

@@

One of the primary limitations of the lineup protocol lies in its
reliance on human judgements. In this context, the effectiveness of a
single lineup evaluation can be dependent on the perceptual ability and
visual skills of the individual. However, when results from multiple
individuals are combined the outcome is encouragingly high-quality and
robust. For simple plots and strong patterns just a few individuals are
needed to arrive at a clear answer, but more individuals will be needed
when the plot design is complex, or the signal strength is weak.
